---
title: "NHANES Sample and Blood Pressure Summary"
author: "Hamed"
date: "2026-10-04"
output:
  pdf_document:
    toc: true
    keep_tex: true
bibliography: ../../references/references.bib
link-citations: true
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  size = "\\scriptsize"
)
options(width = 70)

local_lib <- normalizePath(
  file.path("..", "..", ".r-lib"),
  mustWork = FALSE
)
if (dir.exists(local_lib)) {
  .libPaths(c(local_lib, .libPaths()))
}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
pacman::p_load(tidyverse, knitr, kableExtra, here)

nhanes <- read_csv(
  here("data", "cleaned_NHANES.csv"),
  show_col_types = FALSE
) %>%
  select(-any_of("Unnamed: 0")) %>%
  mutate(
    wave = recode(
      as.character(wave_id),
      `2015` = "2013-2014",
      `2016` = "2015-2016",
      `2017` = "2017-2018"
    ),
    wave = factor(
      wave,
      levels = c("2013-2014", "2015-2016", "2017-2018")
    ),
    readings_available = rowSums(
      !is.na(across(starts_with("systolic_bp_")))
    ),
    avg_sbp = rowMeans(
      across(starts_with("systolic_bp_")),
      na.rm = TRUE
    ),
    avg_sbp = if_else(
      readings_available == 0,
      NA_real_,
      avg_sbp
    )
  )
```

# Introduction

The National Health and Nutrition Examination Survey combines interviews and
health examinations to describe the health of the United States population
[@nhanes]. This report summarizes the three survey waves represented in the
cleaned course dataset. It describes the sample composition and the available
systolic blood-pressure measurements using tables designed for precise
comparison.

# Sample size and male participants

```{r wave-summary}
wave_summary <- nhanes %>%
  group_by(wave) %>%
  summarise(
    sample_size = n(),
    male_n = sum(gender == "Male", na.rm = TRUE),
    male_percent = round(male_n / sample_size * 100, 1),
    .groups = "drop"
  )

largest_wave <- which.max(wave_summary$sample_size)

wave_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Wave", "Sample size", "Male, n", "Male, %"),
    align = c("l", "r", "r", "r"),
    caption = "Overall sample size and male participants by survey wave"
  ) %>%
  kable_styling(
    latex_options = "hold_position",
    full_width = FALSE
  ) %>%
  row_spec(largest_wave, bold = TRUE, background = "lightgray") %>%
  column_spec(1, width = "2.5cm") %>%
  footnote(
    general = "The shaded row has the largest overall sample size.",
    footnote_as_chunk = FALSE
  )
```

The 2013-2014 wave contains the largest sample. Across all three waves, the
numbers of males and females are similar, with males accounting for roughly
half of each wave. Reporting both counts and percentages makes it possible to
compare composition even though the wave sample sizes differ.

# Ethnicity by wave

```{r ethnicity-summary}
ethnicity_summary <- nhanes %>%
  filter(!is.na(ethnicity_2)) %>%
  count(wave, ethnicity_2, name = "n") %>%
  group_by(wave) %>%
  mutate(percent = round(n / sum(n) * 100, 1)) %>%
  ungroup()

wave_rows <- ethnicity_summary %>%
  count(wave) %>%
  tibble::deframe()

largest_ethnicity <- which.max(ethnicity_summary$percent)

ethnicity_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    col.names = c("Wave", "Ethnicity", "n", "%"),
    align = c("l", "l", "r", "r"),
    caption = "Ethnicity counts and percentages within each survey wave"
  ) %>%
  kable_styling(
    latex_options = c("repeat_header"),
    full_width = FALSE
  ) %>%
  pack_rows(index = wave_rows) %>%
  row_spec(largest_ethnicity, bold = TRUE, background = "lightgray") %>%
  footnote(
    general = "Percentages are calculated within wave. The largest displayed percentage is shaded.",
    footnote_as_chunk = FALSE
  )
```

White participants form the largest category in each wave. The table keeps the
more detailed `ethnicity_2` classification because it identifies Asian
participants separately instead of combining them with the broad Other group.
The grouped rows help the reader compare categories within one wave before
moving to the next.

# Average systolic blood pressure

```{r sbp-summary}
sbp_summary <- nhanes %>%
  filter(!is.na(avg_sbp)) %>%
  group_by(wave) %>%
  summarise(
    n = n(),
    minimum = round(min(avg_sbp), 1),
    maximum = round(max(avg_sbp), 1),
    median = round(median(avg_sbp), 1),
    mean = round(mean(avg_sbp), 1),
    sd = round(sd(avg_sbp), 1),
    .groups = "drop"
  )

highest_mean <- which.max(sbp_summary$mean)

sbp_summary %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c(
      "Wave", "Available n", "Minimum", "Maximum",
      "Median", "Mean", "SD"
    ),
    align = c("l", "r", "r", "r", "r", "r", "r"),
    digits = 1,
    caption = "Summary statistics for average systolic blood pressure by wave"
  ) %>%
  kable_styling(
    latex_options = "hold_position",
    full_width = FALSE,
    font_size = 9
  ) %>%
  add_header_above(c(" " = 2, "Average systolic BP (mm Hg)" = 5)) %>%
  row_spec(highest_mean, bold = TRUE, background = "lightgray") %>%
  footnote(
    general = "Average SBP is the participant-level mean of all available systolic readings. The row with the highest mean is shaded.",
    footnote_as_chunk = FALSE
  )
```

The number of participants with an average systolic blood-pressure value is
smaller than the total sample because some participants have no usable
systolic readings. The mean and median are close within each wave, while the
minimum, maximum, and standard deviation show that individual measurements
still vary substantially. Calculating the participant-level average first
prevents participants with more readings from receiving more weight in the
wave summary.

# References

