---
title: "Plotting of NHANES dataset"
author: "Hamed"
date: "2026-10-04"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

local_lib <- normalizePath(
  file.path("..", "..", ".r-lib"),
  mustWork = FALSE
)
if (dir.exists(local_lib)) {
  .libPaths(c(local_lib, .libPaths()))
}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
pacman::p_load(tidyverse, cowplot, ggpubr, DT, here)

hanes <- read_csv(
  here("data", "cleaned_NHANES.csv"),
  show_col_types = FALSE
)

diet <- read_csv(
  here("data", "diet.csv"),
  show_col_types = FALSE
)
```

# Exercise 1 {.tabset}

## Individual figures

```{r exercise-1-individual}
hanes <- hanes %>%
  mutate(
    age_group = cut(age, breaks = seq(15, 90, 5), right = FALSE),
    gender = factor(gender)
  )

p1 <- ggplot(hanes, aes(age_group, fill = gender)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Participants by age and gender",
    x = "Age group (years)",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(hanes, aes(ethnicity_1, fill = gender)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Gender composition by ethnicity",
    x = "Ethnicity",
    y = "Proportion",
    fill = "Gender"
  ) +
  theme_minimal()

p1
p2
```

## Combined figures

```{r exercise-1-combined, fig.height=8}
plot_grid(p1, p2, ncol = 1, align = "v")

ggpubr::ggarrange(
  p1,
  p2,
  ncol = 1,
  common.legend = TRUE,
  legend = "right"
)
```

For these figures, I prefer `plot_grid()` because it gives both plots the same
width while leaving each legend beside its own plot. The vertical arrangement
also gives the long ethnicity labels enough space. `ggarrange()` produces a
cleaner shared legend, but the separate legends make the two mappings easier to
check while I am still developing the figures.

# Exercise 2

The interactive table below makes it possible to search, sort, and page through
the cleaned NHANES data before reviewing the summary figures.

```{r exercise-2-interactive-table}
DT::datatable(
  hanes,
  rownames = FALSE,
  caption = "Cleaned NHANES dataset",
  filter = "top",
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    searching = TRUE,
    ordering = TRUE
  )
)
```

```{r exercise-2-plots}
missing_counts <- hanes %>%
  summarise(
    across(
      c(age, gender, ethnicity_1, ethnicity_2),
      ~ sum(is.na(.))
    )
  )
missing_counts

age_plot <- ggplot(hanes, aes(age)) +
  geom_histogram(
    binwidth = 5,
    fill = "steelblue",
    colour = "white"
  ) +
  labs(
    title = "Age distribution",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()

gender_plot <- ggplot(hanes, aes(gender, fill = gender)) +
  geom_bar() +
  labs(
    title = "Participants by gender",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

e1 <- ggplot(hanes, aes(ethnicity_1, fill = ethnicity_1)) +
  geom_bar() +
  labs(
    title = "Ethnicity variable 1",
    x = "Ethnicity",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 35, hjust = 1)
  )

e2 <- ggplot(hanes, aes(ethnicity_2, fill = ethnicity_2)) +
  geom_bar() +
  labs(
    title = "Ethnicity variable 2",
    x = "Ethnicity",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 35, hjust = 1)
  )

age_plot
gender_plot
e1
e2

ggsave(here("figures", "A_age_distribution_week4.png"), age_plot)
ggsave(here("figures", "A_gender_distribution_week4.png"), gender_plot)
ggsave(here("figures", "A_ethnicity_1_week4.png"), e1)
ggsave(here("figures", "A_ethnicity_2_week4.png"), e2)
```

The age plot shows that the sample includes children, working-age adults, and
older adults, rather than being concentrated in one narrow age range. Gender is
summarized with counts because it is categorical. The missing-value table gives
the exact number of missing observations for all four variables.

I would retain `ethnicity_2`. It separates the Asian category from the broader
"Other" group, so it preserves more information than `ethnicity_1`. I would
remove `ethnicity_1` from the working dataset after documenting this choice.

# Exercise 3

```{r exercise-3}
diet2 <- diet %>%
  group_by(Participant) %>%
  arrange(Week, .by_group = TRUE) %>%
  mutate(baseline = first(Weight)) %>%
  ungroup()

ggplot(diet2, aes(Week, Weight, group = Participant)) +
  geom_line(alpha = 0.2, colour = "grey40") +
  geom_smooth(
    aes(group = 1),
    method = "loess",
    se = FALSE,
    colour = "#0072B2",
    linewidth = 1.2
  ) +
  labs(
    title = "Participant weight generally decreases during the diet",
    subtitle = "Grey lines are participants; blue line is the overall trend",
    x = "Week",
    y = "Weight"
  ) +
  theme_minimal()
```

The original figure assigns a different colour and legend entry to every
participant, creating clutter. Transparency and one overall smooth line
preserve individual trajectories while directing attention to the overall
change.

# Exercise 4

```{r exercise-4}
bp <- hanes %>%
  mutate(
    readings_available = rowSums(
      !is.na(across(starts_with("systolic_bp_")))
    ),
    avg_sbp = rowMeans(
      across(starts_with("systolic_bp_")),
      na.rm = TRUE
    ),
    avg_sbp = if_else(
      readings_available == 0,
      NA_real_,
      avg_sbp
    ),
    sbp_group = case_when(
      avg_sbp < 120 ~ "Normal",
      avg_sbp < 130 ~ "Elevated",
      avg_sbp < 140 ~ "Stage 1",
      TRUE ~ "Stage 2"
    ),
    age_group = cut(
      age,
      breaks = seq(0, 90, by = 5),
      right = FALSE,
      include.lowest = TRUE
    )
  ) %>%
  filter(!is.na(age), !is.na(gender), !is.na(avg_sbp))

ggplot(bp, aes(age_group, fill = sbp_group)) +
  geom_bar(position = "fill") +
  facet_wrap(~ gender) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "YlOrRd") +
  labs(
    title = "Blood-pressure categories by age and gender",
    x = "Age group",
    y = "Proportion",
    fill = "Average SBP"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The complete-case filter means the final distributions describe participants
with age, gender, and at least one systolic blood-pressure reading. The ordered
categories, sequential palette, and faceting make the pattern easier to compare.

